\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{float}
\usepackage{listings}
\usepackage[margin=1.7cm]{geometry}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{hyperref}
\setlength{\parskip}{0.2ex}
\setlength{\parindent}{0pt}
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!10},
  numbers=left,
  numberstyle=\tiny\color{gray},
}
\title{\textbf{Predicting Airbnb Rental Demand: Comparing\\
Logistic Regression, CatBoost, and FT-Transformer with\\
NLP-Engineered Features on Australian Listings}\\[0.4em]
\large COMP90051 Statistical Machine Learning Assignment 2 Group Project}
\author{Kezia Angelina Sutjianta (1576561) \and Anuj Shrivastava (1380251) \and [Student ID] -- [Name]}
\date{}

\begin{document}
\maketitle

%──────────────────────────────────────────────────────────────────────────────
\section{Introduction}
%──────────────────────────────────────────────────────────────────────────────

The short-term rental market has expanded substantially, with platforms such as
Airbnb hosting millions of listings worldwide. Predicting listing demand is
commercially important for hosts, property managers, and platform operators.
Existing work on Airbnb demand prediction largely relies on structured attributes
such as price, location, and review scores \cite{rezazadeh2021}. However,
listings similar in price and review scores often differ in the quality and tone
of their host profiles and text descriptions --- differences that structured
features cannot directly capture.

\textbf{Research question:} \textit{Does NLP-derived text from host profiles and
guest reviews improve demand prediction beyond base listing features alone?}
This goes beyond the obvious prediction task: rather than simply predicting
availability from price and location, we design a controlled ablation where three
classifiers at different complexity tiers are each evaluated on two feature sets
--- base-only versus base plus NLP-engineered text features --- isolating the
marginal value of unstructured text.

\textbf{Dataset:} We use publicly available data from Inside Airbnb
(\url{https://insideairbnb.com}) \cite{insideairbnb}, covering eight Australian
cities (Brisbane, Melbourne, Sunshine Coast, Mornington Peninsula, Northern
Rivers, Barossa Valley, Barwon South West, Mid-North Coast) across eight monthly
snapshots (May~2025--January~2026). After filtering inactive and long-term
listings, the training set comprises 115{,}961 listings across seven snapshots.
Sydney, Tasmania, and Western Australia (31{,}675 listings, latest snapshot) are
withheld as a geographic generalisation split.

%──────────────────────────────────────────────────────────────────────────────
\section{Literature Review}
%──────────────────────────────────────────────────────────────────────────────

Rezazadeh Kalehbasti et al.\ \cite{rezazadeh2021} benchmark machine learning
models for Airbnb price prediction, finding gradient-boosted trees consistently
outperform linear models on structured tabular features. Their work excludes
unstructured text, motivating our NLP ablation design.

Hutto and Gilbert \cite{hutto2014} introduced VADER, a lexicon-based sentiment
analyser calibrated for informal text. We apply VADER to host profiles, listing
descriptions, and guest reviews to extract compound, positive, negative, and
neutral sentiment scores, providing fast text feature engineering without
requiring neural embeddings.

Prokhorenkova et al.\ \cite{prokhorenkova2018} introduced CatBoost, demonstrating
that \textit{ordered boosting} eliminates prediction-shift bias in standard
gradient boosting by computing residuals on a permutation-ordered subset of
training data. This is our medium-complexity baseline and the qualifying
algorithm (NeurIPS 2018).

Gorishniy et al.\ \cite{gorishniy2021} propose the Feature Tokenizer +
Transformer (FT-Transformer), showing Transformer architectures match or exceed
gradient-boosted trees on several tabular benchmarks when each feature is first
embedded via a learned per-feature linear map (NeurIPS 2021). Notably, the paper
acknowledges that tree-based models remain competitive; our results corroborate
this finding on a real-world demand prediction task.

%──────────────────────────────────────────────────────────────────────────────
\section{Methods}
%──────────────────────────────────────────────────────────────────────────────

\subsection{Feature Construction and Preprocessing}

Raw listings from Inside Airbnb provide structured attributes and free-text
fields. We engineer 97 features across four groups (Table~\ref{tab:features}).
All statistics used in feature construction are computed only on the training
portion of each outer fold to prevent leakage.

\begin{table}[H]
\centering\small
\begin{tabular}{p{2.9cm}cp{9.8cm}}
\toprule
Feature group & Count & Description \\
\midrule
Base controls   & 64 &
  Price and price-derived ratios; capacity (accommodates, bedrooms, beds,
  bathrooms); minimum nights; host tenure, response rate, superhost status,
  identity verification, instant bookability; review scores; amenity group
  scores; location features (distance from city centroid). \\[2pt]
Description NLP & 11 &
  VADER compound, positive, negative, neutral sentiment; Flesch-Kincaid grade;
  character count; word count; sentence count; average words per sentence;
  type-token ratio; presence indicator. \\[2pt]
Host profile NLP & 11 &
  Same eleven NLP features applied to the \texttt{host\_about} text field. \\[2pt]
Review NLP      & 11 &
  VADER compound mean, std, min, max across all listing reviews; review count
  before snapshot cutoff; Flesch-Kincaid mean; recency-weighted sentiment;
  positive, negative, and neutral means. \\
\bottomrule
\end{tabular}
\caption{Feature groups. Total: 97 features (64 base + 33 NLP).}
\label{tab:features}
\end{table}

\textbf{Target construction.}
The demand target is binary: a listing is labelled high-demand if its
unavailability rate $r = (90 - \texttt{availability\_90})/90$ exceeds the
75th percentile computed on the training fold (threshold $= 0.733$, positive-class
rate $= 0.25$). We exclude \texttt{estimated\_occupancy\_l365d}: Inside Airbnb
computes it from review counts (Pearson $r = 0.638$ with
\texttt{number\_of\_reviews\_ltm}), creating feature--target leakage when review
NLP features are present. Long-term rentals (\texttt{minimum\_nights} $> 90$)
and never-reviewed listings (\texttt{number\_of\_reviews} $= 0$) are removed
before splitting. Missing numeric values are replaced with per-column medians
computed on each outer training fold.

\subsection{Algorithms}

\begin{table}[H]
\centering\small
\begin{tabular}{llll}
\toprule
Complexity & Algorithm & Tuned parameter(s) & Citation \\
\midrule
Simple  & Logistic Regression (L2) & $C \in \{0.0001, 0.001, 0.01\}$  & --- \\
Medium  & CatBoost                 & depth $\in \{9, 12, 15\}$         & \cite{prokhorenkova2018} \\
Complex & FT-Transformer           & \#blocks $\in \{4, 6, 8\}$        & \cite{gorishniy2021} \\
\bottomrule
\end{tabular}
\caption{Algorithms selected for comparison.}
\label{tab:algorithms}
\end{table}

\textbf{Logistic Regression (simple).}
L2-penalised logistic regression optimises the cross-entropy loss, yielding a
linear decision boundary with feature standardisation applied internally.
$C = 0.01$ was selected in all 10 outer folds. With 115{,}961 training instances
the model benefits from less aggressive regularisation than the earlier 53K-row
pilot, as more data reduces the risk of overfitting.

\textbf{CatBoost (medium, NeurIPS 2018).}
Gradient-boosted symmetric trees with ordered boosting \cite{prokhorenkova2018}.
Standard gradient boosting suffers from prediction-shift bias: the gradient for
a sample is computed using predictions that incorporate that sample. CatBoost
eliminates this via permutation-ordered residuals, yielding unbiased gradients.
Tree depth $= 15$ was selected in all 10 folds; the larger training set affords
deeper trees without overfitting. Fixed: 100 iterations, learning rate $= 0.1$.

\textbf{FT-Transformer (complex, NeurIPS 2021).}
Each feature is mapped to a $d$-dimensional embedding via
$\mathbf{t}_i = x_i\mathbf{W}_i + \mathbf{b}_i$ (Feature Tokenizer)
\cite{gorishniy2021}. A learnable \texttt{[CLS]} token is prepended and
processed by $L$ Pre-LayerNorm Transformer encoder layers with GELU activations
and multi-head self-attention, allowing all feature pairs to interact at every
layer. Implemented from scratch in PyTorch; run on Colab T4 GPU (${\approx}750$s/fold).
Fixed: $d = 32$, 4 heads, feedforward dim $= 128$, AdamW lr $= 3{\times}10^{-4}$,
15 epochs, patience 5.

\subsection{Cross-Validation and Hyperparameter Tuning}

\textbf{Outer CV.} 10-fold cross-validation ($k = 10$), implemented from scratch.
All models use identical folds. The demand-target threshold is recomputed on
each outer training fold to prevent distributional leakage.

\textbf{Inner CV.} Within each outer training fold, hyperparameters are selected
by nested 3-fold CV (also from scratch) using macro-F1 as the inner criterion.
Grid search covers all candidates in Table~\ref{tab:hpgrid}. Logistic and
CatBoost were tuned on 115{,}961 training rows; FT-Transformer was tuned on
the initial 53{,}348 row subset due to GPU runtime constraints.

\begin{table}[H]
\centering\small
\begin{tabular}{lll}
\toprule
Algorithm & Hyperparameter & Candidate values \\
\midrule
Logistic Regression & $C$ (inverse regularisation) & $\{0.0001,\ 0.001,\ 0.01\}$ \\
CatBoost            & Tree depth                   & $\{9,\ 12,\ 15\}$           \\
FT-Transformer      & Encoder layers               & $\{4,\ 6,\ 8\}$             \\
\bottomrule
\end{tabular}
\caption{Hyperparameter grids for nested cross-validation.}
\label{tab:hpgrid}
\end{table}

%──────────────────────────────────────────────────────────────────────────────
\section{Results and Discussion}
%──────────────────────────────────────────────────────────────────────────────

All metrics are computed from scratch. ROC-AUC uses the rank-sum formula;
error bars are $1.96\times\hat\sigma/\sqrt{10}$.
Table~\ref{tab:cv} reports 10-fold CV results; Table~\ref{tab:test} reports
held-out test and geographic generalisation; Table~\ref{tab:cm} shows the
confusion matrix for the best CV configuration.

\begin{table}[H]
\centering\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{llccccc}
\toprule
Model & Features & Accuracy & Macro-F1 & Macro-Prec. & Macro-Rec. & ROC-AUC \\
\midrule
Logistic       & Base only   & $0.838\pm0.001$ & $0.741\pm0.002$ & $0.794\pm0.002$ & $0.700\pm0.002$ & $0.792\pm0.002$ \\
Logistic       & Base+Text   & $0.838\pm0.002$ & $0.742\pm0.002$ & $0.794\pm0.002$ & $0.701\pm0.002$ & $0.794\pm0.002$ \\
FT-Transformer & Base only   & $0.877\pm0.003$ & $0.810\pm0.005$ & $0.844\pm0.005$ & $0.779\pm0.005$ & $0.856\pm0.005$ \\
FT-Transformer & Base+Text   & $0.878\pm0.003$ & $0.810\pm0.004$ & $0.843\pm0.005$ & $0.779\pm0.004$ & $0.855\pm0.004$ \\
CatBoost       & Base only   & $0.885\pm0.002$ & $0.830\pm0.002$ & $0.888\pm0.002$ & $0.790\pm0.002$ & $0.903\pm0.002$ \\
CatBoost       & Base+Text   & $0.887\pm0.002$ & $0.833\pm0.003$ & $0.890\pm0.002$ & $0.793\pm0.003$ & $\mathbf{0.904\pm0.002}$ \\
\bottomrule
\end{tabular}
\caption{10-fold CV results: mean $\pm$ 95\% CI. Logistic and CatBoost trained
  on 115{,}961 listings; FT-Transformer trained on 53{,}348 listings (Colab GPU).}
\label{tab:cv}
\end{table}

\begin{table}[H]
\centering\small
\setlength{\tabcolsep}{5pt}
\begin{tabular}{llcccc}
\toprule
Model & Features &
  \multicolumn{2}{c}{Temporal test ($n=52{,}006$)} &
  \multicolumn{2}{c}{Geographic gen ($n=31{,}675$)} \\
\cmidrule(lr){3-4}\cmidrule(lr){5-6}
 & & Macro-F1 & ROC-AUC & Macro-F1 & ROC-AUC \\
\midrule
Logistic       & Base only   & 0.629 & 0.753 & 0.591 & 0.714 \\
Logistic       & Base+Text   & 0.631 & 0.755 & 0.592 & 0.716 \\
FT-Transformer & Base only   & ---   & ---   & ---   & ---   \\
FT-Transformer & Base+Text   & ---   & ---   & ---   & ---   \\
CatBoost       & Base only   & 0.676 & 0.806 & 0.632 & 0.739 \\
CatBoost       & Base+Text   & \textbf{0.682} & \textbf{0.813} & \textbf{0.639} & \textbf{0.743} \\
\bottomrule
\end{tabular}
\caption{Held-out test set evaluation. Models fit on full training set with best
  CV hyperparameters; threshold from training only.
  FT-Transformer test evaluation deferred due to GPU runtime constraints.}
\label{tab:test}
\end{table}

\begin{table}[H]
\centering\small
\begin{tabular}{llcc}
\toprule
 & & \multicolumn{2}{c}{\textbf{Predicted}} \\
\cmidrule(l){3-4}
 & & Low demand (0) & High demand (1) \\
\midrule
\multirow{2}{*}{\textbf{Actual}}
  & Low demand (0) & 8{,}175 (TN) &   522 (FP) \\
  & High demand (1) &   812 (FN) & 2{,}087 (TP) \\
\bottomrule
\end{tabular}
\caption{Confusion matrix for CatBoost (base only), per outer CV fold average
  ($n_{\text{test}}\approx11{,}596$, positive rate $= 0.25$).
  Recall~$=0.720$, Specificity~$=0.940$.}
\label{tab:cm}
\end{table}

\textbf{Model complexity and CV performance.}
CatBoost (AUC $= 0.903$) substantially outperforms both Logistic Regression
($0.792$, $\Delta = 0.111$) and FT-Transformer ($0.856$, $\Delta = 0.047$).
The large gap over Logistic Regression confirms that short-stay demand is
governed by strong non-linear interactions --- between price, location, capacity,
and review scores --- that gradient-boosted trees capture far more effectively
than a linear boundary. The confusion matrix (Table~\ref{tab:cm}) reveals a
precision--recall trade-off: CatBoost achieves high specificity ($0.940$) but
moderate recall ($0.720$), correctly identifying 70\% of high-demand listings
while producing few false positives. FT-Transformer outperforms Logistic
Regression ($0.856$ vs $0.792$) but falls well below CatBoost, suggesting that
attention-based feature interactions help on this task yet cannot replicate the
efficiency of axis-aligned tree splits over structured demand signals.
Notably, the FT-Transformer's encoder-layer count showed no consistent winner
across folds (4, 6, and 8 each selected 3--4 times), suggesting depth-insensitivity
on this tabular task, consistent with Gorishniy et al.\ \cite{gorishniy2021}.

\textbf{Hyperparameter selection.}
Logistic Regression selected $C = 0.01$ and CatBoost selected depth $= 15$
in all 10 outer folds --- both are the upper bounds of their respective grids
$\{0.0001, 0.001, 0.01\}$ and $\{9, 12, 15\}$. With 115{,}961 training instances
(more than double the earlier 53K pilot), the optimal hyperparameters shifted
toward less regularisation and deeper trees: larger datasets reduce overfitting
risk, favouring higher model capacity. This is a consistent statistical pattern
and an informative finding in itself.

\textbf{Temporal and geographic generalisation.}
Both models show a substantial CV-to-test AUC drop (CatBoost: $0.903 \to 0.806$,
$\Delta = -0.097$; Logistic: $0.792 \to 0.753$, $\Delta = -0.039$). This
reflects \textit{temporal distribution shift}: training used May--November 2025
snapshots while the test set is the January 2026 snapshot, capturing different
seasonal demand patterns. CatBoost is more affected than Logistic because its
deeper trees fit the training distribution more tightly. Geographic generalisation
to Sydney, Tasmania, and Western Australia incurs a further drop (CatBoost:
$0.806 \to 0.739$; Logistic: $0.753 \to 0.714$), reflecting city-specific demand
patterns --- local tourism seasonality, housing stock composition, and regulatory
differences --- not observed during training. Despite both drops, CatBoost
consistently outperforms Logistic Regression on all evaluation sets
($\Delta\text{AUC} \approx 0.05$).

\textbf{Effect of NLP-engineered text features.}
In cross-validation, NLP features provide minimal benefit: $\Delta\text{AUC}
= +0.003$ for Logistic and $+0.001$ for CatBoost. However, on the held-out sets
a small but consistent positive effect emerges: CatBoost gains $+0.008$ AUC on
the temporal test and $+0.003$ on the geographic generalisation; Logistic gains
$+0.002$ and $+0.001$ respectively. This pattern --- near-zero benefit in CV,
small positive effect on unseen data --- suggests text features capture stable
listing quality signals (host engagement, description richness) that complement
structured features specifically under distribution shift, when structured quality
proxies such as review scores and superhost status may behave differently in
new time periods or markets.

\textbf{Alternatives considered.}
TabNet \cite{arik2021} was rejected as the complex model (AAAI does not qualify
under the required venues). LightGBM was rejected for being the same model class
as CatBoost. \texttt{estimated\_occupancy\_l365d} was rejected as a target due
to feature--target leakage ($r = 0.638$ with review counts).

%──────────────────────────────────────────────────────────────────────────────
\section{Conclusion}
%──────────────────────────────────────────────────────────────────────────────

This study compared Logistic Regression, CatBoost (NeurIPS 2018), and
FT-Transformer (NeurIPS 2021) for predicting high-demand Airbnb listings from
97 engineered features across 115{,}961 Australian properties, with and without
NLP-derived text. Three findings emerge. First, CatBoost achieves the best CV
performance (AUC $= 0.903$), outperforming Logistic Regression by $\Delta = 0.111$
and FT-Transformer by $\Delta = 0.047$, confirming that demand is governed by
non-linear structural interactions best captured by gradient-boosted trees.
Second, all models suffer temporal and geographic distribution shift, with AUC
dropping by $0.097$ (CatBoost) and $0.039$ (Logistic) on the January 2026 test
set; a further drop occurs on unseen cities (Sydney, Tasmania, WA). Third, NLP
text features show negligible benefit in cross-validation but a small consistent
positive effect under distribution shift ($\Delta\text{AUC} = +0.003$ to $+0.008$),
suggesting text encodes stable quality signals that complement structured features
when market conditions change.

%──────────────────────────────────────────────────────────────────────────────
\begin{thebibliography}{9}
%──────────────────────────────────────────────────────────────────────────────

\bibitem{insideairbnb}
Cox, M.\ (2025). \textit{Inside Airbnb: Adding data to the debate}.
\url{https://insideairbnb.com}.

\bibitem{rezazadeh2021}
Rezazadeh Kalehbasti, P., Nikolenko, L., \& Rezaei, H.\ (2021).
Airbnb price prediction using machine learning and sentiment analysis.
\textit{Advances in Cross-Cultural Decision Making}, 12, 205--217.
\url{https://arxiv.org/abs/1907.12665}.

\bibitem{hutto2014}
Hutto, C.\ J., \& Gilbert, E.\ (2014).
VADER: A parsimonious rule-based model for sentiment analysis of social media text.
\textit{Proceedings of the 8th AAAI International Conference on Weblogs and
Social Media}, 216--225.

\bibitem{prokhorenkova2018}
Prokhorenkova, L., Gusev, G., Vorobev, A., Dorogush, A.\ V., \& Gulin, A.\
(2018). CatBoost: Unbiased boosting with categorical features.
\textit{Advances in Neural Information Processing Systems}, 31.
\url{https://proceedings.neurips.cc/paper/2018/hash/14491b756b3a51daac41c24863285549-Abstract.html}.

\bibitem{gorishniy2021}
Gorishniy, Y., Rubachev, I., Khrulkov, V., \& Babenko, A.\ (2021).
Revisiting deep learning models for tabular data.
\textit{Advances in Neural Information Processing Systems}, 34, 18932--18943.
\url{https://arxiv.org/abs/2106.11959}.

\bibitem{arik2021}
Arik, S.\ \"{O}., \& Pfister, T.\ (2021).
TabNet: Attentive interpretable tabular learning.
\textit{Proceedings of the AAAI Conference on Artificial Intelligence},
35, 6679--6687.
\url{https://arxiv.org/abs/1908.07442}.

\end{thebibliography}

\end{document}